---
title: "07. Sample Size & Power"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---

## 1. Core Concepts

### The Four Quantities

Sample size and power calculations revolve around four inter-related quantities. Given any three, the fourth is determined:

| Quantity | Symbol | Typical value |
|---|---|---|
| Type I error (significance level) | $\alpha$ | 0.025 (one-sided) |
| Type II error | $\beta$ | 0.10 or 0.20 |
| Power | $1 - \beta$ | 80% or 90% |
| Effect size | $\delta$ | Clinically meaningful difference |

---

### Basic Formula (Two-Sample t-test)

For comparing two means with equal group sizes $n$:

$$n = \frac{(z_{1-\alpha} + z_{1-\beta})^2 \cdot 2\sigma^2}{\delta^2}$$

where $\delta$ is the target difference and $\sigma$ is the standard deviation.

---

## 2. Key Inputs and Assumptions

Getting the assumptions right is critical — a trial powered on wrong assumptions may be severely underpowered.

| Input | Source | Notes |
|---|---|---|
| Effect size $\delta$ | Clinically meaningful difference | Not the "expected" effect — the minimum that matters |
| SD $\sigma$ | Pilot studies, literature, historical data | Often the hardest to estimate |
| Dropout rate | Historical data | Inflate $n$ accordingly |
| $\alpha$ | Pre-specified (regulatory) | One-sided 0.025 is standard |
| Power | Pre-specified | 80% minimum; 90% preferred |

---

## 3. Inflation for Dropout

If dropout proportion is $d$, inflate the calculated $n$ by:

$$n_\text{adj} = \frac{n}{1 - d}$$

For example, with 20% expected dropout and $n = 100$ per arm: $n_\text{adj} = 100 / 0.8 = 125$ per arm.

---

## 4. Sample Size for Common Designs

### Survival Endpoint (Log-rank Test)

For survival trials, the required number of **events** (not subjects) is:

$$E = \frac{4(z_{1-\alpha} + z_{1-\beta})^2}{[\log(\text{HR})]^2}$$

Then sample size is derived from the expected event rate and follow-up time.

### Non-inferiority Trial

The margin $\Delta$ must be pre-specified. The sample size formula becomes:

$$n = \frac{(z_{1-\alpha} + z_{1-\beta})^2 \cdot 2\sigma^2}{(\delta - \Delta)^2}$$

where $\delta$ is the expected treatment difference and $\Delta$ is the NI margin.

---

## 5. Sample Size Re-estimation

Blinded sample size re-estimation (SSR) allows adjustment of $n$ based on observed nuisance parameters (e.g., pooled variance) **without unblinding**. This is permitted by regulators if:

- Pre-specified in the protocol/SAP
- Based only on blinded information (not treatment effect)
- Does not inflate type I error

Unblinded SSR is more complex and requires alpha adjustment.

---

## 6. R Code Practice

```{r}
# install.packages(c("pwr", "gsDesign", "survival", "dplyr", "ggplot2"))
library(pwr)
library(gsDesign)
library(ggplot2)
```

```{r}
# --- Two-sample t-test ---
# Target: detect difference of 5 units, SD = 12, alpha = 0.025 one-sided, power = 90%

result_t <- pwr.t.test(
  d     = 5 / 12,      # Cohen's d = delta / sigma
  sig.level = 0.025,
  power = 0.90,
  type  = "two.sample",
  alternative = "greater"
)
result_t

cat(sprintf("Required n per arm: %d\n", ceiling(result_t$n)))
```

```{r}
# --- Power curve: vary assumed SD ---
sds     <- seq(8, 20, by = 0.5)
n_per_arm <- sapply(sds, function(s) {
  ceiling(pwr.t.test(d = 5 / s, sig.level = 0.025, power = 0.90,
                     type = "two.sample", alternative = "greater")$n)
})

power_curve <- data.frame(sd = sds, n = n_per_arm)

ggplot(power_curve, aes(x = sd, y = n)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_vline(xintercept = 12, linetype = "dashed", color = "red") +
  labs(title = "Required Sample Size vs. Assumed SD",
       subtitle = "Delta = 5, alpha = 0.025 (one-sided), power = 90%",
       x = "Assumed SD", y = "n per arm") +
  theme_minimal()
```

```{r}
# --- Survival endpoint: number of events ---
# Hazard ratio = 0.70, alpha = 0.025 one-sided, power = 90%

gs_surv <- nSurv(
  lambdaC = log(2) / 12,   # Median survival 12 months in control
  hr      = 0.70,           # Target hazard ratio
  sided   = 1,
  alpha   = 0.025,
  beta    = 0.10,
  T       = 36,             # Total study duration (months)
  minfup  = 12              # Minimum follow-up (months)
)

summary(gs_surv)
```

```{r}
# --- Non-inferiority sample size ---
# Expected difference: 0 (new = control)
# NI margin: -3 units
# SD = 10, alpha = 0.025 one-sided, power = 80%

delta_ni  <- 0      # Expected difference (new - control)
margin    <- 3      # NI margin (must not be worse by more than 3)
sd_ni     <- 10
alpha_ni  <- 0.025
power_ni  <- 0.80

z_alpha <- qnorm(1 - alpha_ni)
z_beta  <- qnorm(power_ni)

n_ni <- (z_alpha + z_beta)^2 * 2 * sd_ni^2 / (delta_ni + margin)^2
cat(sprintf("NI trial: required n per arm = %d\n", ceiling(n_ni)))
```

```{r}
# --- Sensitivity: power vs. effect size ---
deltas <- seq(2, 10, by = 0.5)
powers <- sapply(deltas, function(d) {
  pwr.t.test(n = 130, d = d / 12, sig.level = 0.025,
             type = "two.sample", alternative = "greater")$power
})

ggplot(data.frame(delta = deltas, power = powers), aes(x = delta, y = power)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "red") +
  geom_hline(yintercept = 0.90, linetype = "dashed", color = "orange") +
  labs(title = "Power vs. True Effect Size (n = 130/arm)",
       x = "True Difference (delta)", y = "Power") +
  theme_minimal()
```

---

## 7. Key References

- [ICH E9 Statistical Principles for Clinical Trials](https://www.fda.gov/media/71336/download)
- Chow, S.C., Shao, J., & Wang, H. (2008). *Sample Size Calculations in Clinical Research*. CRC Press.
- `pwr` R package: <https://cran.r-project.org/package=pwr>
- `gsDesign` R package: <https://cran.r-project.org/package=gsDesign>
