---
title: "05. Multiple Testing"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---

## 1. Why Multiple Testing Matters

When multiple hypotheses are tested simultaneously, the chance of at least one false positive (type I error) inflates. In clinical trials, this is a serious regulatory concern because an incorrectly approved drug causes real harm.

If each of $m$ tests has type I error $\alpha$, the **familywise error rate (FWER)** under independence is:

$$\text{FWER} = 1 - (1 - \alpha)^m$$

For $m = 5$ and $\alpha = 0.05$: FWER $\approx 0.226$.

---

## 2. Error Rate Definitions

| Rate | Definition | Typical use |
|---|---|---|
| FWER | $P(\text{at least one false positive})$ | Confirmatory trials (strict) |
| FDR | Expected proportion of false positives among rejections | Exploratory, genomics |
| Per-comparison error rate | $\alpha$ per individual test | Usually not acceptable in trials |

Regulatory agencies (FDA, EMA) require **FWER control** at $\alpha = 0.025$ (one-sided) for confirmatory trials.

---

## 3. Common Procedures

### Bonferroni

Reject $H_i$ if $p_i \leq \alpha / m$.

- Simple, conservative
- Does not use correlation structure among tests

### Holm (step-down Bonferroni)

Order p-values: $p_{(1)} \leq p_{(2)} \leq \cdots \leq p_{(m)}$.  
Reject $H_{(i)}$ if $p_{(j)} \leq \alpha / (m - j + 1)$ for all $j \leq i$.

- Uniformly more powerful than Bonferroni
- Still controls FWER strongly

### Hochberg (step-up)

Start from the largest p-value. Reject $H_{(i)}$ and all smaller if $p_{(i)} \leq \alpha / (m - i + 1)$.

- More powerful than Holm under positive dependence
- Requires positive correlation assumption (often satisfied in trials)

### Benjamini-Hochberg (FDR)

Order p-values and reject $H_{(i)}$ if $p_{(i)} \leq \frac{i}{m} \alpha$.

- Controls FDR, not FWER
- Used in exploratory settings, not confirmatory trials

---

## 4. Hierarchical Testing (Fixed Sequence)

A common and simple approach in confirmatory trials: pre-specify a **testing sequence** based on clinical importance. Test in order; proceed to the next hypothesis only if the current one is rejected.

$$H_1 \to H_2 \to H_3$$

- Full $\alpha$ is available at each step
- No $\alpha$ penalty as long as the sequence is pre-specified
- Example: primary endpoint → key secondary → supportive secondary

---

## 5. Graphical Approach (Bretz et al.)

The graphical approach (implemented in `gMCP`) generalizes the weighted Bonferroni procedure. Each hypothesis $H_i$ is assigned:

- An initial weight $w_i$ (proportion of $\alpha$)
- Transition weights $g_{ij}$ (how $\alpha$ flows if $H_i$ is rejected)

When $H_i$ is rejected, its $\alpha$ is redistributed to remaining hypotheses according to $g_{ij}$. This allows flexible, transparent multiplicity strategies that are easy to communicate and pre-specify in SAPs.

---

## 6. R Code Practice

```{r}
# install.packages(c("gMCP", "ggplot2"))
library(gMCP)
library(ggplot2)
```

```{r}
# --- Simple procedures ---

# Simulated p-values from 5 tests
p_vals <- c(0.012, 0.048, 0.021, 0.091, 0.003)
names(p_vals) <- paste0("H", 1:5)

# Bonferroni
p.adjust(p_vals, method = "bonferroni")

# Holm
p.adjust(p_vals, method = "holm")

# Hochberg
p.adjust(p_vals, method = "hochberg")

# Benjamini-Hochberg (FDR)
p.adjust(p_vals, method = "BH")
```

```{r}
# --- Fixed sequence testing ---
# Pre-specified order: H1 -> H2 -> H3
alpha <- 0.025
p_seq <- c(H1 = 0.010, H2 = 0.032, H3 = 0.008)

rejected <- logical(length(p_seq))
for (i in seq_along(p_seq)) {
  if (p_seq[i] <= alpha) {
    rejected[i] <- TRUE
  } else {
    break  # Stop at first non-rejection
  }
}

data.frame(
  Hypothesis = names(p_seq),
  p_value    = p_seq,
  Rejected   = rejected
)
```

```{r}
# --- Graphical approach with gMCP ---
# Two primary endpoints (H1, H2) and two secondary (H3, H4)
# Split alpha equally between H1 and H2
# If H1 rejected -> alpha flows to H3; if H2 -> H4; H3/H4 -> each other

m <- matrix(0, nrow = 4, ncol = 4,
            dimnames = list(paste0("H", 1:4), paste0("H", 1:4)))

# Transition weights (where alpha goes after rejection)
m["H1", "H3"] <- 1     # H1 -> H3
m["H2", "H4"] <- 1     # H2 -> H4
m["H3", "H4"] <- 0.5   # H3 -> H4 (half)
m["H3", "H2"] <- 0.5   # H3 -> H2 (half)
m["H4", "H3"] <- 0.5
m["H4", "H1"] <- 0.5

# Initial alpha weights
weights <- c(H1 = 0.5, H2 = 0.5, H3 = 0, H4 = 0)

graph <- new("graphMCP",
             m = m,
             weights = weights)

plot(graph, main = "Multiplicity Graph")
```

```{r}
# Apply the graph to p-values
p_test <- c(H1 = 0.010, H2 = 0.040, H3 = 0.018, H4 = 0.060)

result <- gMCP(graph, pvalues = p_test, alpha = 0.025)
result
```

---

## 7. Key References

- Bretz, F., Hothorn, T., & Westfall, P. (2010). *Multiple Comparisons Using R*. CRC Press.
- [FDA Guidance on Multiple Endpoints (2022)](https://www.fda.gov/regulatory-information/search-fda-guidance-documents/multiple-endpoints-clinical-trials-guidance-industry)
- `gMCP` R package: <https://cran.r-project.org/package=gMCP>
