---
title: "08. Regulatory & Documentation"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---

## 1. Key ICH Guidelines

The International Council for Harmonisation (ICH) guidelines define the regulatory framework for clinical trials across the US, EU, and Japan.

| Guideline | Topic | Key Points |
|---|---|---|
| ICH E3 | Clinical Study Report (CSR) | Structure and content of the final trial report |
| ICH E6(R2) | Good Clinical Practice (GCP) | Conduct, documentation, monitoring of trials |
| ICH E8 | General Considerations for Clinical Studies | Overall trial design principles |
| ICH E9 | Statistical Principles for Clinical Trials | Primary/secondary endpoints, analysis sets, multiplicity |
| ICH E9(R1) | Estimand Addendum | Estimand framework, intercurrent events |
| ICH E10 | Choice of Control Group | Placebo, active control, NI margins |

---

## 2. Statistical Analysis Plan (SAP)

The SAP is a pre-specified document that describes **exactly how the data will be analyzed** before the blind is broken. It supplements the protocol and is required for regulatory submissions.

### SAP Contents

- **Analysis populations**: FAS, PP, safety set — definitions and handling of deviations
- **Primary analysis**: exact model, covariates, estimand, handling of missing data
- **Secondary analyses**: listed in testing hierarchy
- **Multiplicity strategy**: testing procedure, graphical approach
- **Subgroup analyses**: pre-specified vs. exploratory
- **Sensitivity analyses**: alternative models, MNAR scenarios
- **Tables, Listings, Figures (TLFs)**: mock shells pre-specified

### Timeline

| Milestone | SAP Status |
|---|---|
| Protocol finalized | Draft SAP begins |
| Database lock imminent | SAP must be finalized and signed |
| After unblinding | SAP amendments require justification |

The SAP must be **finalized before unblinding**. Any post-unblinding changes must be fully disclosed.

---

## 3. Analysis Populations

| Set | Definition | Used for |
|---|---|---|
| ITT (Intent-to-Treat) | All randomized patients | Principle, not always primary |
| FAS (Full Analysis Set) | All randomized, received ≥1 dose, ≥1 post-baseline measurement | Primary efficacy |
| PP (Per-Protocol) | Completed trial without major deviations | Sensitivity analysis |
| Safety Set | All who received ≥1 dose | Safety analyses |
| mITT | Subset of ITT after removing patients with data quality issues | Sometimes used as primary |

---

## 4. CDISC Standards

CDISC (Clinical Data Interchange Standards Consortium) provides standardized data formats required for FDA submissions.

### SDTM (Study Data Tabulation Model)

Raw collected data, organized by domain:

| Domain | Content |
|---|---|
| DM | Demographics |
| AE | Adverse events |
| LB | Laboratory results |
| VS | Vital signs |
| CM | Concomitant medications |
| EX | Exposure (dosing) |

### ADaM (Analysis Data Model)

Derived analysis-ready datasets:

| Dataset | Content |
|---|---|
| ADSL | Subject-level baseline and treatment info |
| ADTTE | Time-to-event analysis data |
| ADLB | Analysis laboratory data |
| ADAE | Analysis adverse event data |
| ADEFF | Efficacy analysis data (trial-specific) |

Key ADaM variables:

- `PARAMCD` / `PARAM`: parameter code and label
- `AVAL`: analysis value
- `BASE`: baseline value
- `CHG` / `PCHG`: change / percent change from baseline
- `ANL01FL`: analysis flag (which records to include in primary analysis)

---

## 5. Clinical Study Report (ICH E3)

The CSR is the comprehensive report submitted to regulators after trial completion. Key sections:

| Section | Content |
|---|---|
| Synopsis | One-page summary of objectives, design, results |
| Introduction | Background and rationale |
| Study objectives | Primary and secondary endpoints, estimands |
| Investigational plan | Design, randomization, blinding, dosing |
| Study patients | Disposition, deviations, analysis populations |
| Efficacy evaluation | Primary and secondary results, subgroups |
| Safety evaluation | AEs, SAEs, labs, vital signs, deaths |
| Statistical methods | Reference to SAP |
| Discussion and conclusions | Interpretation |
| Appendices | Protocol, SAP, informed consent, TLFs |

---

## 6. R Code Practice

```{r}
# install.packages(c("pharmaverseadam", "dplyr", "ggplot2", "gtsummary"))
library(pharmaverseadam)
library(dplyr)
library(ggplot2)
library(gtsummary)
```

```{r}
# Load CDISC pilot ADSL dataset
data(adsl)
glimpse(adsl)
```

```{r}
# Table 1: Baseline characteristics by treatment arm (SAP-style)
adsl |>
  select(TRT01P, AGE, SEX, RACE, BMIBL, WEIGHTBL) |>
  tbl_summary(
    by = TRT01P,
    label = list(
      AGE      ~ "Age (years)",
      SEX      ~ "Sex",
      RACE     ~ "Race",
      BMIBL    ~ "BMI at baseline",
      WEIGHTBL ~ "Weight at baseline (kg)"
    ),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    )
  ) |>
  add_overall() |>
  bold_labels()
```

```{r}
# Analysis population flags: how many in FAS vs. PP?
adsl |>
  count(SAFFL, ITTFL, EFFFL) |>
  arrange(desc(n))
```

```{r}
# Load ADTTE for survival analysis on CDISC pilot data
data(adtte)
glimpse(adtte)

# Event counts by treatment
adtte |>
  filter(PARAMCD == "TTDE") |>   # Time to first dermatological event
  group_by(TRT01P) |>
  summarise(
    n_subjects = n(),
    n_events   = sum(CNSR == 0),
    event_rate = mean(CNSR == 0)
  )
```

```{r}
# Disposition summary (common in CSR Section 10)
adsl |>
  count(TRT01P, DCREASCD) |>
  filter(!is.na(DCREASCD)) |>
  ggplot(aes(x = DCREASCD, y = n, fill = TRT01P)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title  = "Reason for Discontinuation by Treatment Arm",
    x      = NULL, y = "Number of Subjects",
    fill   = "Treatment"
  ) +
  theme_minimal()
```

---

## 7. Key References

- [ICH E9 Statistical Principles](https://www.fda.gov/media/71336/download)
- [ICH E9(R1) Estimand Addendum](https://www.ich.org/page/efficacy-guidelines)
- [ICH E3 Clinical Study Report](https://www.ich.org/page/efficacy-guidelines)
- [CDISC Standards](https://www.cdisc.org/standards)
- `pharmaverseadam` R package: <https://cran.r-project.org/package=pharmaverseadam>
