---
title: "06. Interim Analysis"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---

## 1. What is Interim Analysis?

An interim analysis is a pre-planned examination of accumulating trial data **before the final analysis**. It allows a trial to be stopped early for:

- **Efficacy**: The treatment effect is already convincing
- **Futility**: The treatment is unlikely to succeed even if continued
- **Safety**: Unacceptable harm is observed

Because interim looks at the data inflate type I error, special statistical methods are needed to maintain the overall $\alpha$ level.

---

## 2. The Multiple Looks Problem

If you test at $\alpha = 0.05$ at both an interim and final analysis, the true FWER is not 0.05. Under the null hypothesis, testing twice inflates error:

$$P(\text{reject at least once} \mid H_0) > \alpha$$

The solution is to **spend** $\alpha$ carefully across looks.

---

## 3. Alpha Spending Functions

The **alpha spending function** $\alpha^*(t)$ determines how much type I error to spend at information fraction $t \in [0, 1]$, where $t = 1$ is the final analysis.

Requirements:
- $\alpha^*(0) = 0$
- $\alpha^*(1) = \alpha$
- $\alpha^*(t)$ is non-decreasing

### O'Brien-Fleming (OBF)

Spends very little $\alpha$ early — requires a very strong effect to stop early.

$$\alpha^*_\text{OBF}(t) = 2\left(1 - \Phi\left(\frac{z_{\alpha/2}}{\sqrt{t}}\right)\right)$$

- Conservative early stopping boundary
- Most commonly used in confirmatory trials

### Pocock

Spends $\alpha$ evenly across looks — easier to stop early but reduces power at the final analysis.

- Less popular in confirmatory settings

### Lan-DeMets

Generalizes OBF and Pocock as continuous functions of information fraction $t$. Does not require pre-specifying the number or timing of interim looks in advance.

---

## 4. Boundaries

At each interim look $k$, the test statistic $Z_k$ is compared to a boundary $c_k$:

- **Efficacy boundary**: Stop and reject $H_0$ if $Z_k \geq c_k$
- **Futility boundary**: Stop and fail to reject if $Z_k \leq f_k$ (binding or non-binding)

Non-binding futility boundaries do not affect the type I error calculation — the trial can continue past them without penalty.

---

## 5. DSMB (Data Safety Monitoring Board)

A DSMB (or IDMC) is an **independent committee** that reviews unblinded interim data and recommends whether to continue, modify, or stop the trial. Key features:

- Members are independent from the sponsor
- Operates under a **charter** that pre-specifies the monitoring plan
- Maintains blinding of the sponsor team
- Makes recommendations; the sponsor makes the final decision

The statistical team prepares a **closed report** (unblinded) for the DSMB and an **open report** (blinded) for the sponsor.

---

## 6. Conditional Power and Futility

**Conditional power** is the probability of achieving a significant result at the final analysis, given current data and assuming the original treatment effect.

$$\text{CP}(t) = P\left(Z_\text{final} \geq c_\text{final} \mid Z_t, \delta = \hat{\delta}\right)$$

If conditional power drops below a threshold (e.g., 20%), the trial may be stopped for futility.

---

## 7. R Code Practice

```{r}
# install.packages("gsDesign")
library(gsDesign)
library(ggplot2)
```

```{r}
# --- Two-look group sequential design ---
# 1 interim + 1 final, O'Brien-Fleming spending, one-sided alpha = 0.025

gs_design <- gsDesign(
  k      = 2,          # Number of analyses (1 interim + 1 final)
  test.type = 1,       # One-sided
  alpha  = 0.025,
  beta   = 0.1,        # 90% power
  sfu    = sfLDOF,     # Lan-DeMets O'Brien-Fleming spending
  delta  = 1           # Standardized effect size (will be normalized)
)

summary(gs_design)
```

```{r}
# Plot the design: efficacy and futility boundaries
plot(gs_design, main = "Group Sequential Design: OBF Spending")
```

```{r}
# --- Three-look design with futility ---
gs3 <- gsDesign(
  k         = 3,
  test.type = 4,       # Two-sided with non-binding futility
  alpha     = 0.025,
  beta      = 0.1,
  sfu       = sfLDOF,
  sfl       = sfLDPocock,  # Futility spending
  delta     = 1
)

summary(gs3)
plot(gs3)
```

```{r}
# --- Alpha spending: compare OBF vs. Pocock ---
t_vals <- seq(0, 1, by = 0.01)

obf     <- sfLDOF(alpha = 0.025, t = t_vals)$spend
pocock  <- sfLDPocock(alpha = 0.025, t = t_vals)$spend

spend_df <- data.frame(
  t       = rep(t_vals, 2),
  spend   = c(obf, pocock),
  method  = rep(c("OBF", "Pocock"), each = length(t_vals))
)

ggplot(spend_df, aes(x = t, y = spend, color = method)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Alpha Spending Functions",
    x = "Information Fraction (t)",
    y = "Cumulative Alpha Spent",
    color = "Method"
  ) +
  theme_minimal()
```

```{r}
# --- Conditional power calculation ---
# Assume: z-statistic at interim = 1.8, information fraction = 0.5
# Original target effect: delta = 0.3 (on standardized scale)

z_interim   <- 1.8
t_interim   <- 0.5
z_final_crit <- 1.96  # Two-sided 0.05 (or OBF boundary at final)

# Conditional power (assuming original effect holds)
# Z_final | Z_t ~ N(z_interim + delta * sqrt(1 - t), 1) approximately
delta_std <- 0.3
z_cp <- (z_final_crit - z_interim - delta_std * sqrt(1 - t_interim)) /
        sqrt(1 - t_interim)

cp <- 1 - pnorm(z_cp)
cat(sprintf("Conditional power: %.1f%%\n", cp * 100))
```

---

## 8. Key References

- Lan, K.K.G. & DeMets, D.L. (1983). Discrete sequential boundaries for clinical trials. *Biometrika*.
- Jennison, C. & Turnbull, B.W. (2000). *Group Sequential Methods with Applications to Clinical Trials*. CRC Press.
- [FDA Guidance on Adaptive Design (2019)](https://www.fda.gov/regulatory-information/search-fda-guidance-documents/adaptive-designs-clinical-trials-drugs-and-biologics)
- `gsDesign` R package: <https://cran.r-project.org/package=gsDesign>
