---
title: "04. Missing Data"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---

## 1. Missing Data Mechanisms

Understanding *why* data are missing determines which methods are valid.

### MCAR — Missing Completely At Random

The probability of missingness is unrelated to any observed or unobserved data.

$$P(R = 0 \mid Y_\text{obs}, Y_\text{mis}) = P(R = 0)$$

Example: A blood sample is lost in the lab due to a processing error.

- Complete case analysis is **unbiased** (but inefficient)
- Rare in clinical trials

---

### MAR — Missing At Random

The probability of missingness depends only on **observed** data, not unobserved outcomes.

$$P(R = 0 \mid Y_\text{obs}, Y_\text{mis}) = P(R = 0 \mid Y_\text{obs})$$

Example: Older patients drop out more, and age is recorded. Conditional on age, dropout is random.

- Most standard methods (MMRM, multiple imputation) are valid under MAR
- The most common assumption in regulatory submissions

---

### MNAR — Missing Not At Random

The probability of missingness depends on the **unobserved** outcome itself.

$$P(R = 0 \mid Y_\text{obs}, Y_\text{mis}) \neq P(R = 0 \mid Y_\text{obs})$$

Example: Patients drop out because their condition worsened — and that worsening is the missing value.

- Standard methods are **biased** under MNAR
- Requires sensitivity analyses (tipping point, pattern-mixture models)

---

## 2. Common Methods

| Method | Valid under | Notes |
|---|---|---|
| Complete case analysis | MCAR | Simple but loses data |
| LOCF (last obs. carried forward) | Neither | Biased under MAR; avoid as primary |
| MMRM | MAR | Preferred for continuous outcomes |
| Multiple imputation (MI) | MAR | Flexible; works for various outcomes |
| Pattern-mixture models | MNAR (assumption explicit) | Sensitivity analysis |
| Tipping point analysis | MNAR | How bad must MNAR be to overturn results? |

ICH E9(R1) recommends that the **primary analysis target the estimand** and handle missing data in a principled way aligned with the intercurrent event strategy.

---

## 3. Multiple Imputation (MI)

MI proceeds in three steps:

1. **Impute**: Generate $m$ complete datasets by drawing from $P(Y_\text{mis} \mid Y_\text{obs})$
2. **Analyze**: Run the analysis model on each of the $m$ datasets
3. **Pool**: Combine estimates using Rubin's rules

### Rubin's Rules

Let $\hat{\theta}_k$ and $\hat{V}_k$ be the estimate and variance from imputation $k$:

$$\bar{\theta} = \frac{1}{m} \sum_{k=1}^m \hat{\theta}_k$$

$$T = \bar{V} + \left(1 + \frac{1}{m}\right) B$$

where $\bar{V}$ is the within-imputation variance and $B = \frac{1}{m-1}\sum(\hat{\theta}_k - \bar{\theta})^2$ is the between-imputation variance.

---

## 4. Sensitivity Analysis under MNAR

After a MAR-based primary analysis, regulators expect sensitivity analyses to assess robustness to MNAR departures.

### Reference-Based Imputation

Impute missing post-dropout data using the trajectory of a **reference group** (e.g., placebo):

- **Jump to reference (J2R)**: After dropout, the patient's outcome immediately follows the reference arm
- **Copy reference (CR)**: The patient's trajectory copies the reference arm from dropout onward
- **Last mean carried forward (LMCF)**: Impute using the last observed mean of the reference group

### Tipping Point Analysis

Find the value of the MNAR departure parameter $\delta$ at which the treatment effect becomes non-significant. If the tipping point is implausibly large, the conclusion is robust.

---

## 5. R Code Practice

```{r}
# install.packages(c("mice", "dplyr", "ggplot2"))
library(mice)
library(dplyr)
library(ggplot2)
```

```{r}
# Use nhanes dataset from mice (small, illustrative)
data(nhanes, package = "mice")
head(nhanes)
summary(nhanes)

# Check missing data pattern
md.pattern(nhanes, rotate.names = TRUE)
```

```{r}
# Multiple imputation with m = 20 imputations
imp <- mice(nhanes, m = 20, method = "pmm", seed = 123, printFlag = FALSE)

# Inspect one completed dataset
complete(imp, 1) |> head()
```

```{r}
# Analyze: linear regression on each imputed dataset
# Outcome: chl (cholesterol), predictor: age + bmi + hyp
fit_list <- with(imp, lm(chl ~ age + bmi + hyp))

# Pool results using Rubin's rules
pooled <- pool(fit_list)
summary(pooled)
```

```{r}
# Visualize: compare observed vs. imputed distributions for bmi
imp_long <- complete(imp, action = "long", include = TRUE)

ggplot(imp_long, aes(x = bmi, fill = factor(.imp == 0))) +
  geom_density(alpha = 0.4) +
  scale_fill_manual(values = c("steelblue", "tomato"),
                    labels = c("Imputed", "Observed")) +
  labs(title = "BMI: Observed vs. Imputed Distributions",
       x = "BMI", fill = "") +
  theme_minimal()
```

```{r}
# Simple tipping point illustration
# Suppose primary analysis gives: estimate = 3.2, SE = 1.1
# Sensitivity: shift imputed values for treatment arm by delta
# Find delta where p-value crosses 0.05

est0 <- 3.2
se0  <- 1.1

deltas  <- seq(0, -5, by = -0.1)
p_vals  <- 2 * pnorm(abs((est0 + deltas) / se0), lower.tail = FALSE)

tipping <- data.frame(delta = deltas, p_value = p_vals)

ggplot(tipping, aes(x = delta, y = p_value)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red") +
  labs(title = "Tipping Point Analysis",
       x = expression(delta ~ "(MNAR shift for treatment arm)"),
       y = "p-value") +
  theme_minimal()
```

---

## 6. Key References

- Rubin, D.B. (1987). *Multiple Imputation for Nonresponse in Surveys*. Wiley.
- [ICH E9(R1) Addendum on Estimands](https://www.ich.org/page/efficacy-guidelines)
- van Buuren, S. (2018). *Flexible Imputation of Missing Data*. <https://stefvanbuuren.name/fimd/>
- `mice` R package: <https://cran.r-project.org/package=mice>
