---
title: "03. MMRM (Mixed Model for Repeated Measures)"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---

## 1. What is MMRM?

MMRM (Mixed Model for Repeated Measures) is the standard approach for analyzing **longitudinal continuous outcomes** in clinical trials. It models all post-baseline measurements jointly, accounting for within-subject correlation across visits.

The model is:

$$Y_{ij} = \mathbf{x}_{ij}^\top \boldsymbol{\beta} + \varepsilon_{ij}$$

where $\boldsymbol{\varepsilon}_i \sim \mathcal{N}(\mathbf{0}, \boldsymbol{\Sigma})$ and $\boldsymbol{\Sigma}$ is typically an unstructured covariance matrix across visits.

Key points:

- **No random effects** — this is a marginal (population-averaged) model
- **Unstructured covariance** is the default — no assumption on how correlations decay over time
- **Valid under MAR** — missing data due to dropout is handled without explicit imputation
- Primary target: **treatment difference at the final visit**

---

## 2. MMRM vs. ANCOVA

| Feature | ANCOVA | MMRM |
|---|---|---|
| Data used | Last visit only (often with LOCF) | All post-baseline visits |
| Dropout handling | Requires imputation | Valid under MAR |
| Statistical efficiency | Lower | Higher |
| Regulatory trend | Older standard | Current preference |

ANCOVA on observed completers or LOCF-imputed data is biased under MAR. MMRM avoids this by using all available data within a likelihood-based framework.

---

## 3. Covariance Structures

| Structure | Description | Notes |
|---|---|---|
| Unstructured (UN) | All entries free | Default; most flexible |
| Compound symmetry (CS) | Equal variance, equal correlation | Assumes exchangeable visits |
| AR(1) | Correlation decays exponentially with lag | Evenly spaced visits |
| Toeplitz | Correlation depends only on lag | More flexible than AR(1) |

For confirmatory trials, **unstructured** is preferred to avoid misspecification. Simpler structures are used when convergence fails or sample sizes are very small.

---

## 4. Connection to Estimands (ICH E9(R1))

MMRM naturally aligns with:

- **Treatment policy strategy**: uses data from patients who discontinue treatment (if collected post-discontinuation)
- **Hypothetical strategy**: models the outcome as if treatment continued

The estimand must be specified first — then the model is chosen to target it.

---

## 5. R Code Practice

```{r}
# Install if needed:
# install.packages(c("mmrm", "dplyr", "ggplot2"))

library(mmrm)
library(dplyr)
library(ggplot2)

# Use the built-in fev_data from mmrm package
# Outcome: FEV1 (lung function), Treatment: TRT (active vs. placebo)
# Visits: AVISIT, Subject: USUBJID, Baseline: BASE

data(fev_data, package = "mmrm")
head(fev_data)
```

```{r}
# Fit MMRM with unstructured covariance
# FEV1 ~ treatment * visit + baseline + (unstructured by visit | subject)

fit <- mmrm(
  formula = FEV1 ~ RACE + SEX + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data
)

summary(fit)
```

```{r}
# Extract treatment contrasts at each visit
library(emmeans)

emm <- emmeans(fit, ~ ARMCD | AVISIT)
contrast(emm, method = "pairwise")
```

```{r}
# Plot observed means by treatment group and visit
fev_data |>
  group_by(ARMCD, AVISIT) |>
  summarise(mean_fev = mean(FEV1, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = AVISIT, y = mean_fev, color = ARMCD, group = ARMCD)) +
  geom_line() +
  geom_point(size = 2) +
  labs(
    title = "Observed Mean FEV1 by Visit and Treatment",
    x = "Visit", y = "Mean FEV1", color = "Treatment"
  ) +
  theme_minimal()
```

```{r}
# Compare covariance structures: unstructured vs. AR(1)
fit_ar1 <- mmrm(
  formula = FEV1 ~ RACE + SEX + ARMCD * AVISIT + ar1(AVISIT | USUBJID),
  data = fev_data
)

# AIC comparison
AIC(fit, fit_ar1)
```

---

## 6. Key References

- [ICH E9(R1) Addendum on Estimands](https://www.ich.org/page/efficacy-guidelines)
- Mallinckrodt et al. (2008). *Recommendations for the Primary Analysis of Continuous Endpoints in Longitudinal Clinical Trials*. Drug Information Journal.
- `mmrm` R package documentation: <https://openpharma.github.io/mmrm/>
