---
title: "01. Estimand Framework"
subtitle: "ICH E9(R1) & Analysis Populations in Clinical Trials"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---

## Overview

The estimand framework was introduced in ICH E9(R1) (2019) to address a fundamental
ambiguity in clinical trial analysis: the research question being asked and the
statistical analysis performed were often not precisely aligned. The framework
forces explicit agreement on *what* is being estimated before choosing *how* to
estimate it.

| Component | Description |
|-----------|-------------|
| Population | Who is being studied |
| Variable | The outcome of interest |
| Intercurrent event strategy | How post-randomization events are handled |
| Population-level summary | The measure of treatment effect (e.g., mean difference, HR) |

---

## 1. What is an Estimand?

An estimand specifies exactly what a clinical trial is trying to estimate.

Historically, stating "primary analysis will follow ITT" left unanswered:

- How to handle patients who switched treatment mid-trial
- How to treat outcomes for patients who died before the endpoint
- Whether rescue medication use should be considered a failure

The estimand framework resolves this by separating the *question* from the *method*.

### The Four Components

```
Estimand = Population
         + Variable (endpoint)
         + Intercurrent event handling strategy
         + Population-level summary
```

**Example:**

> "In patients aged ≥18 with moderate asthma (Population),  
> the change from baseline in FEV1 at week 52 (Variable),  
> regardless of treatment discontinuation (ICE: treatment policy strategy),  
> estimated as the mean difference versus placebo (Summary)"

---

## 2. Intercurrent Events (ICE)

An intercurrent event is any post-randomization event that affects either the
interpretation or existence of the endpoint measurement.

**Common examples:**

- Treatment discontinuation
- Use of rescue medication
- Death (when the endpoint is non-fatal)
- Protocol deviation affecting treatment exposure

### The Five ICE Handling Strategies

| Strategy | Definition | Typical use case |
|----------|-----------|-----------------|
| **Treatment policy** | Analyze as randomized, regardless of ICE | When the real-world effect matters most |
| **Hypothetical** | Estimate the outcome had the ICE not occurred | Isolating the biological effect of treatment |
| **Composite** | Incorporate the ICE into the outcome (e.g., discontinuation = failure) | When the ICE is itself a clinically meaningful event |
| **While on treatment** | Use only data collected prior to the ICE | Pharmacodynamic or mechanism studies |
| **Principal stratum** | Restrict to patients who would not experience the ICE | Subgroup-specific causal estimands |

::: {.callout-note}
Treatment policy strategy and ITT are related but distinct concepts.
ITT defines *who is included* (the analysis population).
Treatment policy defines *how ICEs are handled* (the estimand strategy).
Treatment policy + FAS is the modern equivalent of the traditional ITT analysis.
:::

---

## 3. Analysis Populations

### Definitions

| Population | Inclusion Criteria | Primary use |
|-----------|-------------------|-------------|
| **Full Analysis Set (FAS)** | All randomized patients who received at least one dose | Effectiveness; regulatory submissions |
| **mITT** | Study-specific (e.g., patients with a baseline measurement) | Defined per protocol |
| **Per-Protocol (PP)** | Patients who complied with the protocol without major deviations | Efficacy under ideal conditions |
| **Safety Set** | All patients who received at least one dose | Safety analyses |

### Choosing the Primary Population

```
Superiority trial     → FAS primary, PP supportive
Non-inferiority trial → BOTH FAS AND PP as co-primary
Regulatory submission → FAS/ITT primary
```

::: {.callout-warning}
In non-inferiority trials, FAS alone is insufficient. Non-compliant patients dilute
the difference between arms, which can spuriously support a NI conclusion.
Both FAS and PP must demonstrate non-inferiority.
:::

---

## 4. R Code

```{r}
#| label: setup
library(dplyr)
library(ggplot2)
library(knitr)
```

### 4.1 Simulating Analysis Population Assignment

```{r}
#| label: analysis-sets

set.seed(42)
n <- 200

patients <- tibble(
  id = 1:n,
  randomized         = TRUE,
  received_treatment = sample(c(TRUE, FALSE), n, replace = TRUE, prob = c(0.97, 0.03)),
  protocol_deviated  = sample(c(TRUE, FALSE), n, replace = TRUE, prob = c(0.15, 0.85)),
  baseline_measured  = sample(c(TRUE, FALSE), n, replace = TRUE, prob = c(0.98, 0.02))
) |>
  mutate(
    in_FAS    = randomized & received_treatment,
    in_mITT   = in_FAS & baseline_measured,
    in_PP     = in_FAS & !protocol_deviated & baseline_measured,
    in_safety = received_treatment
  )

patients |>
  summarise(
    FAS    = sum(in_FAS),
    mITT   = sum(in_mITT),
    PP     = sum(in_PP),
    Safety = sum(in_safety)
  ) |>
  kable(caption = "Patient counts by analysis population (simulated)")
```

### 4.2 Effect of ICE Strategy on Estimated Treatment Effect

```{r}
#| label: ice-simulation

set.seed(123)
n_patients <- 100

trial_data <- tibble(
  id           = 1:n_patients,
  arm          = rep(c("Treatment", "Placebo"), each = n_patients / 2),
  week_12      = rnorm(n_patients,
                       mean = ifelse(rep(c("Treatment", "Placebo"), each = 50) == "Treatment", -2, -0.5),
                       sd = 1.5),
  discontinued = sample(c(TRUE, FALSE), n_patients, replace = TRUE, prob = c(0.2, 0.8))
)

composite_data <- trial_data |>
  mutate(week_12_composite = ifelse(discontinued, NA_real_, week_12))

results <- tibble(
  Strategy = c("Treatment Policy", "While on Treatment", "Composite (discontinued = missing)"),
  N = c(
    sum(trial_data$arm == "Treatment"),
    sum(trial_data$arm == "Treatment" & !trial_data$discontinued),
    sum(trial_data$arm == "Treatment")
  ),
  Mean_change = c(
    mean(trial_data$week_12[trial_data$arm == "Treatment"]),
    mean(trial_data$week_12[trial_data$arm == "Treatment" & !trial_data$discontinued]),
    mean(composite_data$week_12_composite[composite_data$arm == "Treatment"], na.rm = TRUE)
  )
)

kable(results, digits = 3,
      caption = "Mean week-12 change by ICE handling strategy (treatment arm)")
```

```{r}
#| label: ice-plot
#| fig-cap: "Same data, different estimand — different estimate"
#| fig-width: 7
#| fig-height: 4

ggplot(results, aes(x = Strategy, y = Mean_change, fill = Strategy)) +
  geom_col(width = 0.5, alpha = 0.85) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  scale_fill_manual(values = c("#2E86AB", "#A23B72", "#F18F01")) +
  labs(
    title = "Estimated treatment effect varies by ICE handling strategy",
    x = NULL,
    y = "Mean change at week 12",
    caption = "Illustrates why pre-specifying the estimand matters"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"))
```

---

## Notes

- The distinction between estimand (what) and estimator (how) is important.
  MMRM, multiple imputation, and last observation carried forward are all
  *estimators* — they answer different estimands even when applied to the same data.
- The choice of ICE strategy should be driven by the clinical question, not
  by what is computationally convenient.
- Sensitivity analyses assess how robust conclusions are to the choice of
  estimand and its underlying assumptions.

---

## Glossary

**Estimand**
: The precise quantity that a clinical trial is designed to estimate. Defined by four
components: population, variable, intercurrent event strategy, and population-level summary.

**Estimator**
: The statistical method used to estimate the estimand (e.g., MMRM, multiple imputation).
Different estimators can be applied to the same estimand, and the same estimator can
be used to answer different estimands.

**Estimate**
: The numerical result produced by applying an estimator to the observed data (e.g., −1.2 units).

**Hazard Ratio (HR)**
: A measure of treatment effect used in time-to-event analyses. Defined as the ratio
of the hazard rate in the treatment arm to the hazard rate in the control arm.
HR < 1 indicates lower risk of the event in the treatment group. Covered in detail in 02-survival-analysis.

**Intent-to-Treat (ITT)**
: The principle of analyzing all randomized patients in the group they were assigned to,
regardless of whether they actually received the treatment or complied with the protocol.
Reflects real-world effectiveness rather than ideal efficacy.

**Full Analysis Set (FAS)**
: The modern implementation of the ITT principle. Includes all randomized patients
who received at least one dose of study treatment.

**Per-Protocol (PP)**
: An analysis population restricted to patients who followed the protocol as planned,
without major deviations. Reflects efficacy under ideal conditions.

**Intercurrent Event (ICE)**
: Any post-randomization event that affects the interpretation or existence of the
endpoint measurement (e.g., treatment discontinuation, rescue medication use, death).

**Randomization**
: The process of randomly assigning participants to treatment or control arms.
Ensures comparability between groups and is the basis for causal inference in clinical trials.

**Effectiveness vs. Efficacy**
: Effectiveness refers to how well a treatment works in real-world conditions (ITT/FAS).
Efficacy refers to how well it works under ideal, controlled conditions (PP).

---

## References

- ICH E9(R1) Addendum on Estimands and Sensitivity Analysis in Clinical Trials (2019)
- Lawrance et al. (2020). *What is an estimand & how does it relate to quantifying
  the effect of treatment on patient-relevant outcomes?*
  Therapeutic Innovation & Regulatory Science.
- Mallinckrodt et al. (2020). *Estimands: Four Essential Questions.*
  Pharmaceutical Statistics.
