---
title: "02. Survival Analysis"
subtitle: "Time-to-Event Methods in Clinical Trials"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---

## Overview

Survival analysis is a class of statistical methods for analyzing the time until
an event of interest occurs. Despite the name, the "event" does not have to be
death — it can be disease progression, hospitalization, relapse, or any
clinically meaningful endpoint.

| Method | Purpose |
|--------|---------|
| Kaplan-Meier (KM) estimator | Non-parametric estimate of the survival function |
| Log-rank test | Compare survival curves between two or more groups |
| Cox proportional hazards model | Estimate the effect of covariates on survival |

---

## 1. Key Concepts

### 1.1 Time-to-Event

The outcome in survival analysis has two components:

- **Time**: duration from a defined starting point (e.g., randomization) to the event
- **Event indicator**: whether the event was observed (1) or the observation was censored (0)

```r
Surv(time, event)   # the basic survival object in R
```

### 1.2 Censoring

Censoring occurs when the event has not been observed by the end of the study.
This is not missing data — it is partial information. The patient was event-free
up to their last observed time point.

**Types of censoring:**

| Type | Definition | Example |
|------|-----------|---------|
| **Right censoring** | Event has not occurred by end of follow-up | Patient still alive at study end |
| **Left censoring** | Event occurred before the observation started | Rarely encountered in trials |
| **Interval censoring** | Event occurred within a known time window | Detected at a scheduled visit |

Right censoring is by far the most common in clinical trials.

**Reasons for right censoring:**

- Study ended before the patient experienced the event
- Patient withdrew consent
- Patient was lost to follow-up

::: {.callout-warning}
A key assumption is that censoring is **non-informative** — the reason a patient
is censored is unrelated to their prognosis. If sicker patients are more likely
to drop out, this assumption is violated and results may be biased.
:::

### 1.3 Hazard and Hazard Ratio

The **hazard** is the instantaneous rate of the event occurring at time t,
given that the patient has survived up to time t.

```
h(t) = lim  P(t ≤ T < t + Δt | T ≥ t) / Δt
       Δt→0
```

The **Hazard Ratio (HR)** compares the hazard between two groups:

```
HR = h_treatment(t) / h_control(t)

HR < 1  →  lower risk of event in treatment group (favorable)
HR = 1  →  no difference between groups
HR > 1  →  higher risk of event in treatment group
```

---

## 2. Kaplan-Meier Estimator

The KM estimator is a non-parametric method for estimating the survival function
S(t) = P(T > t) — the probability of surviving beyond time t.

### How it works

At each event time, the survival probability is updated:

```
S(t) = ∏ (1 - d_i / n_i)

where:
  d_i = number of events at time t_i
  n_i = number of patients at risk just before t_i
```

### Properties

- Makes no assumptions about the shape of the survival distribution
- Handles censoring naturally — censored patients are removed from the risk set
  at their censoring time
- Produces a step function that drops at each observed event time

---

## 3. Log-Rank Test

The log-rank test is the standard non-parametric test for comparing survival
curves between two or more groups. It tests the null hypothesis that the
survival functions are identical across groups.

### How it works

At each event time, the test compares observed vs. expected number of events
in each group, summed across all event times.

### When the log-rank test is NOT appropriate

- When the **proportional hazards assumption is violated** — i.e., when the
  hazard ratio is not constant over time (e.g., survival curves cross)
- When there are **time-varying treatment effects**
- When comparing more than two groups without adjustment for multiplicity

In these cases, alternatives include the **weighted log-rank test**
(e.g., Fleming-Harrington) or **restricted mean survival time (RMST)**.

---

## 4. Cox Proportional Hazards Model

The Cox model is a semi-parametric regression model that estimates the effect
of covariates on the hazard function:

```
h(t | X) = h_0(t) · exp(β_1 X_1 + β_2 X_2 + ... + β_p X_p)

where:
  h_0(t) = baseline hazard (unspecified — this is the non-parametric part)
  exp(β)  = hazard ratio for a one-unit increase in X
```

### Why "semi-parametric"?

- The baseline hazard h_0(t) is left completely unspecified (non-parametric)
- The covariate effects are modeled parametrically via exp(Xβ)
- This makes the Cox model very flexible — no distributional assumption needed

### 4.1 The Proportional Hazards Assumption

The key assumption of the Cox model is that the hazard ratio between any two
patients is **constant over time**. This means the survival curves should not cross.

**How to verify the PH assumption:**

| Method | How |
|--------|-----|
| **Schoenfeld residuals** | Plot residuals vs. time — should show no trend |
| **cox.zph() in R** | Formal test; p < 0.05 suggests violation |
| **Log-log plot** | Plot log(-log(S(t))) vs. log(t) — lines should be parallel |
| **Visual inspection** | KM curves should not cross |

**What to do if PH assumption is violated:**

- Use **time-varying coefficients** in the Cox model
- Use **stratified Cox model** (stratify by the violating variable)
- Switch to **RMST** (restricted mean survival time) as the summary measure
- Use **accelerated failure time (AFT)** models

---

## 5. Competing Risks

Competing risks arise when patients can experience multiple types of events,
and the occurrence of one event prevents or alters the probability of the other.

**Example:** In a cancer trial measuring time to disease progression:
- Disease progression is the event of interest
- Death without progression is a competing risk

Using standard KM in the presence of competing risks **overestimates** the
cumulative incidence of the event of interest, because it treats competing
events as censored observations.

**Solution:** Use the **cause-specific hazard** or the
**cumulative incidence function (CIF)** via the Fine-Gray model.

---

## 6. R Code

```{r}
#| label: setup
library(survival)
library(survminer)
library(dplyr)
library(ggplot2)
```

### 6.1 Simulating Trial Data

```{r}
#| label: simulate-data

set.seed(42)
n <- 200

trial_data <- tibble(
  id        = 1:n,
  arm       = rep(c("Treatment", "Placebo"), each = n / 2),
  time      = rexp(n, rate = ifelse(rep(c("Treatment", "Placebo"), each = n/2) == "Treatment", 0.05, 0.08)),
  event     = rbinom(n, 1, prob = 0.75),
  age       = rnorm(n, mean = 60, sd = 10),
  sex       = sample(c("M", "F"), n, replace = TRUE)
) |>
  mutate(time = pmin(time, 24))  # administrative censoring at 24 months
```

### 6.2 Kaplan-Meier Curve

```{r}
#| label: km-plot
#| fig-cap: "Kaplan-Meier survival curves by treatment arm"
#| fig-width: 8
#| fig-height: 5

km_fit <- survfit(Surv(time, event) ~ arm, data = trial_data)

ggsurvplot(
  km_fit,
  data          = trial_data,
  pval          = TRUE,
  conf.int      = TRUE,
  risk.table    = TRUE,
  legend.labs   = c("Placebo", "Treatment"),
  palette       = c("#E7B800", "#2E86AB"),
  xlab          = "Time (months)",
  ylab          = "Survival probability",
  title         = "Kaplan-Meier Survival Curves",
  ggtheme       = theme_minimal()
)
```

### 6.3 Log-Rank Test

```{r}
#| label: log-rank

logrank_test <- survdiff(Surv(time, event) ~ arm, data = trial_data)
print(logrank_test)
```

### 6.4 Cox Proportional Hazards Model

```{r}
#| label: cox-model

cox_fit <- coxph(Surv(time, event) ~ arm + age + sex, data = trial_data)
summary(cox_fit)
```

### 6.5 Checking the Proportional Hazards Assumption

```{r}
#| label: ph-assumption
#| fig-cap: "Schoenfeld residuals — no trend indicates PH assumption holds"
#| fig-width: 8
#| fig-height: 5

ph_test <- cox.zph(cox_fit)
print(ph_test)   # p > 0.05 → no evidence of PH violation

ggcoxzph(ph_test)
```

### 6.6 Log-Log Plot

```{r}
#| label: log-log-plot
#| fig-cap: "Log-log plot — parallel lines indicate proportional hazards"
#| fig-width: 7
#| fig-height: 5

plot(
  km_fit,
  fun     = "cloglog",
  col     = c("#E7B800", "#2E86AB"),
  xlab    = "log(Time)",
  ylab    = "log(-log(S(t)))",
  main    = "Log-Log Plot for PH Assumption Check"
)
legend("topleft", legend = c("Placebo", "Treatment"),
       col = c("#E7B800", "#2E86AB"), lty = 1)
```

---

## Notes

- The KM estimator is descriptive — it estimates the survival function but does
  not test hypotheses or adjust for covariates.
- The log-rank test is the default test for comparing survival curves, but it
  has low power when the PH assumption is violated or when treatment effects
  emerge late (e.g., immunotherapy trials).
- The Cox model is the workhorse of survival analysis in clinical trials. It
  gives adjusted hazard ratios while making minimal distributional assumptions.
- Always check the PH assumption before interpreting Cox model results.
  A significant cox.zph() test (p < 0.05) warrants further investigation.

---

## Glossary

**Survival function S(t)**
: The probability that the event of interest has not occurred by time t.
Ranges from 1 (at t = 0) to 0 (as t → ∞).

**Hazard function h(t)**
: The instantaneous rate of the event occurring at time t, given survival up to t.
Intuitively, the "risk" of the event at a given moment.

**Hazard Ratio (HR)**
: The ratio of hazard rates between two groups. The primary effect measure in
Cox regression. HR = 0.7 means the treatment group has 30% lower risk of the
event at any point in time (under the PH assumption).

**Censoring**
: Incomplete observation of the event time. A censored patient was event-free
up to their last contact. Not the same as missing data.

**Kaplan-Meier (KM) estimator**
: A non-parametric estimator of the survival function. Makes no assumptions
about the underlying distribution of event times.

**Log-rank test**
: A non-parametric test for comparing survival curves. Most powerful when the
hazard ratio is constant over time (PH assumption holds).

**Cox proportional hazards model**
: A semi-parametric regression model relating covariates to the hazard function.
The most widely used survival model in clinical trials.

**Proportional hazards (PH) assumption**
: The assumption that the hazard ratio between groups is constant over time.
Required for valid interpretation of Cox model coefficients.

**Competing risks**
: A situation where multiple event types exist and the occurrence of one
prevents or modifies the probability of another.

**Restricted Mean Survival Time (RMST)**
: The expected event-free survival time up to a pre-specified time horizon.
A useful alternative to HR when the PH assumption is violated.

---

## References

- Cox, D.R. (1972). *Regression models and life tables.* Journal of the Royal
  Statistical Society, Series B.
- Kaplan, E.L. & Meier, P. (1958). *Nonparametric estimation from incomplete
  observations.* Journal of the American Statistical Association.
- Therneau, T. & Grambsch, P. (2000). *Modeling Survival Data: Extending the
  Cox Model.* Springer.
- Royston, P. & Parmar, M.K.B. (2011). *The use of restricted mean survival time
  to estimate the treatment effect in randomized clinical trials when the
  proportional hazards assumption is in doubt.* Statistics in Medicine.
